---
title: 'N-Gram Text Prediction Model '
author: "Oli Bailey"
date: "September, Year Of The Virus"
output:
  html_document:
    theme: sandstone
  pdf_document: default
subtitle: 'Step 1. Clean the three input texts'
---


## NOTES:

This script conerns itself with the reading in to RAM of the very large "corpus" of three input files, and cleaning these texts before we create the ngram tables in the next file.

The output files are based on a 25% sample of the entire input corpus.

```{r, load_libs, message=FALSE, warning=FALSE}
library(dplyr)      # for wrangling
library(magrittr)   # for piping
library(qdap)       # for "GradyAugmented" dictionary check to remove non-english words
library(stringi)    # for sentence extraction
library(stringr)    # for cleaning

# remove any previous variables (clean start)
rm(list = ls())
gc()
```

```{r, utility_functions}
# file handling
load_file <- function(file_name) {
  fp <- file(file_name)
  f <- readLines(fp)
  # remove non-ASCII characters
  f <- iconv(f, "UTF-8", "ASCII")
  close(fp)
  return(f)
}

# Check a single word
data("GradyAugmented")
is.word  <- function(x) x %in% c(GradyAugmented)
```

Load ALL files  to RAM (~500MB of R object memory)
```{r, load_files, warning=FALSE, message=FALSE}
# we assume the input files are in ./_InputCorpus/ folder relative to this file
# please modify this to suit your folder layout...
blogs <- load_file("../_InputCorpus/en_US.blogs.txt")
news <- load_file("../_InputCorpus/en_US.news.txt")
twitter <- load_file("../_InputCorpus/en_US.twitter.txt")
```

We will base our prediction model on a 25% sample of these files.
```{r, sample_mixture}
desired_percentage <- 25  # of total object sizes in sample

set.seed(123)
reduction_factor <-  desired_percentage / 100
blogs <- sample(blogs,ceiling(length(blogs) * reduction_factor))

set.seed(345)
reduction_factor <-  desired_percentage / 100
news <- sample(news,ceiling(length(news) * reduction_factor))

set.seed(678)
reduction_factor <-  desired_percentage / 100
twitter <- sample(twitter,ceiling(length(twitter) *reduction_factor))

# create combined sample corpus
corpus <- c(blogs,news,twitter)

# release memory
rm (blogs, news, twitter)
gc()
```

Now we want to "clean" the corpus, removing sentences containing foul words, removing misspelled words, and handling both American-English spelling and British-English e.g. color/colour, favorite/favourite.
```{r process_corpus}
# one sentence per row
corpus <- unlist(stri_split_boundaries(corpus, type="sentence"))
corpus <- tibble(text = corpus)

# all words to lower case
# remove most common foul words (ignoring the effect on remaining meaning) 
# remove punctuation (EXCEPT ' and - e.g. "i'm double-barelled")
con = file("http://www.bannedwordlist.com/lists/swearWords.txt")
banned_words = c("motherfucker", readLines(con))
close(con)

# convert to lower case, remove punctuation (except hyphens and apostrophes)
corpus %<>%
    mutate(text = tolower(text)) %>%
    mutate(text = str_remove_all(text, '[#~0-9\\.,><?!"£\\@$%\\^&\\*_;:\\(\\)]')) %>%
    mutate(text = stri_replace_all(str=text, 
                                   replacement = "unkwn", 
                                   regex= paste(banned_words, collapse = "|"))) 

# remove NA lines
corpus <- corpus[!is.na(corpus)]

# add line number into corpus
corpus <- tibble(line_no=1:length(corpus), text = corpus)

# split corpus to words 
# NOTE: this would convert to lower case and removes punctuation if we hadn't done so already!
words <- corpus %>% tidytext::unnest_tokens(text, text)

# remove misspelled words (dictionary check) [NOT RECOMMENDED! SEE NOTES]
#words <- words[is.word(words$text),]

# rebuild corpus 
corpus <- words %>% 
    group_by(line_no) %>% 
    summarize(text = str_c(text, collapse = " ")) %>%
    ungroup()

# remove sentences of only one word (they have no use in our model!)
corpus %<>% filter(stri_count_words(text) > 1)

# remove a 2% chunk of our corpus to use for testing
set.seed(2021)
inTest <- sample(1:nrow(corpus), floor(nrow(corpus)*0.02))
test_corpus <- corpus[inTest,]
corpus <- corpus[-inTest,]

# remove another small chunk (1%) of the corpus as our "held-out"
# corpus for tuning lambda interpolation hyperparameter (TBD: FUTURE REVISION)
inHeldOut <-sample(1:nrow(corpus), floor(nrow(corpus)*0.05))
held_out_corpus <- corpus[inHeldOut,]
corpus <- corpus[-inHeldOut,]


# save training, held-out and test corpora
save(corpus, file="../Data/train_corpus_25.rdat")
save(test_corpus, file="../Data/test_corpus_25.rdat")
save(held_out_corpus, file="../Data/held_out_corpus_25.rdat")

```

So now we have three files, with "corpus" containing most of the data, to be used
to create the ngram tables in the file "create_ngram_tables.Rmd"
